---
title: "Identify the problem and make a plan"
teaching: 50
exercises: 20
---


::: questions
-   What do I do when I encounter an error?
-   What do I do when my code outputs something I don’t expect?
-   Why do errors and warnings appear in R?
-   How can I find which areas of code are responsible for errors?
-   How can I fix my code? What other options exist if I can't fix it?
:::


::: objectives
After completing this episode, participants should be able to...

-   Describe how the desired code output differs from the actual output
-   Categorize an error message (e.g. syntax error, semantic errors, package-specific errors, etc.)
-   decode/describe what an error message is trying to communicate
-   Identify specific lines and/or functions generating the error message
-   Use R Documentation to look up function syntax and examples
-   Quickly fix commonly-encountered R errors using the internet
-   Identify when a problem is better suited for asking for further help, including online help and reprex
:::


```{r, echo=F, message=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(stringr)

# Read in the data
rodents <- read_csv("data/surveys_complete_77_89.csv")

```


(initial intro -- edit upon looking at intro episode)

The first step we'll cover is what to do when encountering an error or other undesired output from your code. With this episode, we hope to teach you the basics about identifying errors, rectifying them if possible, and if not, how to isolate the problem for others to look at. This is the first step in our "roadmap" of how to solve coding problems -- recognizing when something is happening that you don't intend with your code, and then identifying the problem (to a lesser or greater degree) in order to solve it yourself or be able to succintely describe it to a helper. 

## 3.1 What do I do when I encounter an error message?

While sometimes frusturating to read, R will often let us know when a problem occurs by generating an error message that is meant to tell us why R was unable to run our code. This type of ‘error’ is often referred to as a **syntax** error. When R is unable to run your code, it will return this type of error message, and stop the program (as opposed to a **warning** or attempting to run further lines despite the error). Error messages may happen for many reasons. However, deciphering the meaning of such error messages is not always as easy as we might hope. While we can't review every type of reason your code generates an error, we will try to teach you some tools for you to interpret and figure out syntax errors for yourself.

The accompanying error message attempts to tell you exactly how your code failed. For example, consider the following error message that occurs when I run this command in the R console:
  
```{r,error=T}
ggplot(x = taxa) + geom_bar()
```

Though we know somewhere there is an object called `taxa` (it is actually a column of the dataset `rodents`), R is trying to communicate that it cannot find any such object in the local environment. Let's try again, appropriately pointing ggplot to the `rodents` dataset and `taxa` column using the `$` operator. For the sake of argument, let's say we also remember that `geom_bar` expects an aesthetic (`aes`).

```{r, error=T}
ggplot(aes(x = rodents$taxa)) + geom_bar()
```

Whoops! Here we see another error message -- this time, R responds with a perhaps more-uninterpretable message.



Let's go over each part briefly. First, we see an error from a function called `fortify`, which we didn't even call! Then, there's a more helpful informational message: Did we accidentally pass `aes()` to the `data` argument? This does seem to relate to our function call, as we do pass `aes` into the `ggplot` function. But what is this “`data` argument?" A helpful starting place when attempting to decipher an error message is checking the documentation for the function which caused the error:

`?ggplot`

Here, a Help window pops up in RStudio which provides some more information. Skipping the general description at the top, we see ggplot takes positional arguments of `data`, *then* `mapping`, which uses the `aes` call. We can see in "Arguments" that the `aes(x = rodents$taxa)` object used in the plot is attempted by `fortify` to be converted to a data.frame: now the picture is clear! We accidentally passed our `mapping` argument (telling ggplot how to map variables to the plot) into the position it expected `data` in the form of a data frame. And if we scroll down to "Examples", to "Pattern 1", we can see exactly how ggplot expects these arguments in practice. Let's amend our result:
  
```{r}
ggplot(rodents, aes(x = taxa)) + geom_bar()
```

Here we see our desired plot.

## Stop no. 1 on our roadmap: Identifying the problem 

Let's pause here to highlight some patterns we're starting to see in the course of fixing our code: 

1. Seeing a problem arise in our code (in this case, R is explicitly telling us it has a problem running it).

1. Reading and interpreting the error message R gives us. 

Other steps we might take then include: 

3. Acting on parts of the error we can understand, such as changing input to a function. 

3. Pulling up the R Documentation for that function, and reading the documentation's Description, Usage, Arguments, Details and Examples entries for greater insight into our error.

3. Copying and pasting the error message into a search engine / generative LLM for more interpretable explanations.

And, when all else fails, we can prepare our code into a reproducible example for expert help.

While the above steps may be new or seem familiar, we formalize this a little bit to explicitly address the first two points: recognizing when a problem arises and attempting to interpret *what* is going wrong is essential to fixing it. This is true whether you fix the problem on your own, or by effectively communicating it to an expert. The latter steps one might attempt to immediately address the problem we'll call **code first aid** -- these steps might fix the problem, give you greater insight into what the problem is (and how R is interpreting your code), or not be helpful at all. 

In any case, we want to emphasize the first skills are essential to being a practiced coder able to effectively seek help. While they may seem trivial, we will see below examples of problems that are trickier to both recognize and interpret. 


## 3.2 What do I do when my code outputs something I don't expect

Another type of problem you may encounter is when your R code runs without errors, but does not produce the desired output. You may sometimes see these called **semantic errors**. As with syntax errors, semantic errors may occur for a variety of non-intuitive reasons, and are often harder to solve as there is no description of the error -- you must work out where your code is defective yourself!
  
Let's go back to our rodent analysis. The next step in the plan is to subset to just the `Rodent` taxa (as opposed to other taxa: Bird, Rabbit, Reptile or NA). Let's quickly check to see how much data we'd be throwing out by doing so:
  
```{r}
table(rodents$taxa)
```

We're interested in the Rodents, and thankfully it seems like a majority of our observations will be maintained when subsetting to rodents. But wait. In our plot above, we can clearly see the presence of NA values. Why are we not seeing them here? This is an example of a semantic error -- our command is executed correctly, but the output is not everything we intended. Having no error message to interpret, let's jump straight to the function documentation:
  
```{r}
?table
```

Here, the documentation provides some clues: `table` takes an argument called `useNA` that accepts "no", "ifany", and "always", but it's not immediately apparent which one we should use to show our NA values. As a second approach, let's go to `Examples` to see if we can find any quick fixes. Here we see a couple lines further down:
  
``` r
table(a)                 # does not report NA's
table(a, exclude = NULL) # reports NA's
```

That seems like it should be inclusive. Let's try again:

```{r}
table(rodents$taxa, exclude = NULL)
```

Now, we do see that by subsetting to the "Rodent" taxa, we are losing about 357 NAs, which themselves could be rodents! However, in this case, it seems a small enough portion to safely omit. Let's subset our data to the rodent taxa

```{r}
rodents <- rodents %>% filter(taxa == "Rodent")
```



::::::: challenge 

There are 3 lines of code below, and each attempts to create the same plot. Identify which produces a syntax error, which produces a semantic error, and which correctly creates the plot (hint: this may require you inferring what type of graph we're trying to create!)




A.  `ggplot(rodents) + geom_bin_2d(aes(month, plot_type))`

B.  `ggplot(rodents) + geom_tile(aes(month, plot_type), stat = "count")`

C.  `ggplot(rodents) + geom_tile(aes(month, plot_type))`


::::: solution

In this case, A correctly creates the graph, plotting as colors in the tile the number of times an observation is seen. It essentially runs the following lines of code:

```{r}
rodents_summary <- rodents %>% group_by(plot_type, month) %>% summarize(count=n())
ggplot(rodents_summary) + geom_tile(aes(month, plot_type, fill=count))
```

B is a syntax error, and will produce the following error: 

```{r, error=T}
ggplot(rodents) + geom_tile(aes(month, plot_type), stat = "count")
```

Finally, C is a semantic error. It produces a plot, which is rather meaningless:

```{r}
ggplot(rodents) + geom_tile(aes(month, plot_type))
```

:::::


::::::: 



## Summary

In general, encountering semantic errors can make our job more difficult, but the roadmap remains the same: 


1. Seeing a problem arise in our code.

1. Interpreting the problem.  

Other steps we might take then include: 

3. Acting on parts of the error we can understand, such as changing input to a function. 

3. Pulling up the R Documentation for that function, and reading the documentation's Description, Usage, Arguments, Details and Examples entries for greater insight into our error.

3. Describing our problem into a search engine / generative LLM for more interpretable explanations.

And, when all else fails, we can prepare our code into a reproducible example for expert help.

The steps to identifying the problem and in *code first aid* matches much of what we've seen above. However, here seeing the problem arise in our code may be much more subtle, and comes from us recognizing output we don't expect or know to be wrong. Even if the code is run, R may give us *warning* or informational messages which pop up when executing your code. Most of the time, however, it's up to the coder to be vigilant and be sure steps are running as they should. Interpreting the problem may also be more difficult as R gives us little or no indication about how it's misinterpreting our intent. 


identifying problems more difficult, as R will not explicitly tell you there's a problem. Interpreting the problem  for which a remedy is not immediately apparent, some steps to take include:
  
1.  Reading any warning or informational messages that may pop up when executing your code.

2.  Pulling up the R Documentation for that function (which may be specific to a package, such as with ggplot).

3.  Reading through the documentation's Description, Usage, Arguments, Details and Examples entries for greater insight into our error.

4.  Changing the input to your function call to see if the output or semantic error can be made more apparent (maybe delete).


And, when all else fails, preparing our code into a reproducible example for expert help. Note, there are fewer options available with semantic errors as when an error message prevents your code from running. You may find yourself having to isolate or ask for help more often than with easily-solvable syntax errors.

::: callout
Generally, the more your code deviates from just using base R functions, or the more you use specific packages, both the quality of documentation and online help available from search engines and Googling gets worse and worse. While base R errors will often be solvable in a couple of minutes from a quick `?help` check or a long online discussion and solutions on a website like Stack Overflow, errors arising from little-used packages applied in bespoke analyses might merit isolating your specific problem to a reproducible example for online help, or even getting in touch with the developers! Such community input and questions are often the way packages and documentation improves over time.
:::

## 3.3 How can I find where my code is failing?

Isolating your problem may not be as simple as assessing the output from a single function call on the line of code which produces your error. Often, it may be difficult to determine which line(s) in your code are producing the error.

Consider the example below, where we now are attempting to see how which species of kangaroo rodents appear in different plot types over the years. To do so, we'll filter our dataset to just include the genus *Dipodomys*. Then we'll plot a histogram of which how many observations are seen in each plot type over an x axis of years.

```{r, error=T}
krats <- rodents %>% filter(genus == "Dipadomys") #kangaroo rat genus

ggplot(krats, aes(year, fill=plot_type)) + 
geom_histogram() +
facet_wrap(~species)

```

Uh-oh. Another error here, this time in "combine_vars?" What is that? "Faceting variables must have at least one value": What does that mean?

Well it may be clear enough that we seem to be missing "species" values where we intend. Maybe we can try to make a different graph looking at what species values are present? Or perhaps there's an error earlier -- our safest approach may actually be seeing what krats looks like:
  
```{r}
krats
```

It's empty! What went wrong with our "Dipadomys" filter? Let's use a print statement to see which genuses are included in the original rodents dataset.

```{r}
print(rodents %>% count(genus))
```

We see two things here. For one, we've misspelled *Dipodomys*, which we can now amend. This quick function call also tells us we should expect a data frame with 9573 values resulting after subsetting to the genus *Dipodomys*.

```{r}
krats <- rodents %>% filter(genus == "Dipodomys") #kangaroo rat genus
dim(krats)

ggplot(krats, aes(year, fill=plot_type)) + 
geom_histogram() +
facet_wrap(~species)

```

Our improved code here looks good. Checking the dimensions of our subsetted data frame using the dim() function confirms we now have all the *Dipodomys* observations, and our plot is looking better. In general, having a 'print' statement or some other output before plots or other major steps can be a good way to check your code is producing intermediate results consistent with your expectations.




However, there's something funky happening here. The bins are definitely weirdly spaced -- we can see some bins are not filled with any observations, while those exactly containing one of the integer years happens to contain all the observations for that year.

A solution here might be to tinker with the bin width in the histogram code, but let's step back a minute. Do we necessarily need to dive into the depths of tinkering with the plot? We can evalulate this problem not in terms of the plot having a problem, but with our data type having a problem. There's an opportunity to encode the observation times outside of coarse, somewhat arbitrary year groupings with the real, interpretable date they were collected. Let's do that using the tidyverse's 'lubridate' package. The important details here are that we are creating a 'datetime'-type variable using the recorded years, months, and days, which are currently all encoded as numeric types.


```{r}
krats <- rodents %>% filter(genus == "Dipodomys") #kangaroo rat genus
dim(krats)

krats <- krats %>% mutate(date = lubridate::ymd(paste(year,month,day,sep='-')))

ggplot(krats, aes(date, fill=plot_type)) + 
  geom_histogram() +
  facet_wrap(~species)

```

This looks much better, and is easier to see the trends over time as well. Note our x axis still shows bins with year labelings, but the continuous spread of our data over these bins shows that dates are treated more continuously and fall more continuously within histogram bins.


One aspect we can see with this exercise above is that sometimes the 'problem' we need to solve is not super intuitive or can be isolated to one line. Before attempting to hack a bespoke solution, sometimes it's helpful to consider whether this might be an often-encountered problem (even if you can't exactly articulate the problem yourself) or one where you have some intuition there is a simpler or more elegant solution. 

By setting up a reproducible example, we can isolate the problem with real data rather than simply asking a proximal question online or to a friend (e.g. 'how can i change my plot to look like so'). This allows the community to assist in identifying the problem -- you don't always need to have it figured out before asking for help! 


:::::: callout
Often, giving your expert helpers access to the entire problem, with a detailed description of your desired output allows you to directly improve your coding skills and learn about new functions and techniques.
:::::: 


## Summary

In general, we need to isolate the specific areas of code causing the bug or problem. There is no general rule of thumb as to how large this needs to be, but in general, think about what we would want to include in a reprex. Any early lines which we know run correctly and as intended may not need to be included, and we should seek to isolate the problem area as much as we can to make it understandable to others.

Let's add to our list of steps for identifying the problem:
  
0.  Identify the problem area -- add print statements immediately upstream or downstream of problem areas, check the desired output from functions, and see whether any intermediate output can be further isolated.

1.  Read each part of the error or warning message (if applicable), to see if we can immediately interpret and act on any.

2.  Pull up R Documentation for any function calls causing the error (which may be specific to a package, such as with ggplot).

3.  Read through the documentation's Description, Usage, Arguments, Details and Examples entries for greater insight into our error.

4.  Copy and paste the error message into a search engine for more interpretable explanations.

And, when all else fails, prepare our code into a reproducible example for expert help.


While this is similar to our previous checklists, we can now understand these steps as a continuous cycle of isolating the problem into more and more discrete chunks for a reproducible example. Any step in the above that helps us identify the specific areas or aspects of our code that are failing in particular, we can zoom in on and restart the checklist. We can stop as soon as we don't understand anymore how our code fails. At this point, we can excise that area for further help using a reprex.

Finally, let's make our plot publication-ready by changing some aesthetics. Let's also add a vertical line to show when the study design changed on the exclosures.

```{r}
krats <- rodents %>% filter(genus == "Dipodomys") #kangaroo rat genus
dim(krats)

krats <- krats %>% mutate(date = lubridate::ymd(paste(year,month,day,sep='-')))


krats %>%
  ggplot(aes(x = date, fill = plot_type)) +
  geom_histogram()+
  facet_wrap(~species, ncol = 1)+ 
  theme_bw()+
  scale_fill_viridis_d(option = "plasma")+
  geom_vline(aes(xintercept = lubridate::ymd("1988-01-01")), col = "dodgerblue")

```

It looks like the study change helped to reduce *merriami* sightings in the Rodent and Short-term Krat exclosures.




While it is common practice in intro coding courses to call over the instructor with a raised hand and a statement such as "I don't know what's wrong," in reality people have limited time, bandwidth, or requisite knowledge to be able to help out with any problem that might arise. Even if they can't figure out a bug on their own, the practiced coder can identify and articulate the problem effectively such that someone with available time and expertise can help out 

